### 2023-10-13 draft_jss/draft_jss.Rmd
rename jss

---
documentclass: jss
author:
  - name: FirstName LastName
    orcid: 0000-0000-0000-0000
    affiliation: University/Company
    # use this syntax to add text on several lines
    address: |
      | First line
      | Second line
    email: \email{name@company.com}
    url: https://posit.co
  - name: Second Author
    orcid: 0000-0000-0000-0000
    affiliation: 'Affiliation \AND'
    # To add another line, use \AND at the end of the previous one as above
  - name: Third Author
    orcid: 0000-0000-0000-0000
    address: |
      | Department of Statistics and Mathematics,
      | Faculty of Biosciences,
      | Universitat Autònoma de Barcelona
    affiliation: |
      | Universitat Autònoma 
      | de Barcelona
    # use a different affiliation in adress field (differently formated here)
    affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "A Capitalized Title: Something about a Package \\pkg{foo}"
  # If you use tex in the formatted title, also supply version without
  plain:     "A Capitalized Title: Something about a Package foo"
  # For running headers, if needed
  short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
---

@@

1. Initialize :
  - Population parameters : Npop = 200, Ne = 100, Ngeneration = 30
  - Cross-over parameters : Ntournament = 2 
  - Mutation parameters : pmutQuant = .5, pmutCat = .25, sigma = 1 (0.1 for leaking rate as the hyperparameter range is shorter)
  - Define the hyperparameter_list, a list containing each param.
  - For each param in hyperparameter_list define hyperparameter search_space : define the lower and upper bound, the type (numeric, integer, categorical) and if it should be log transformed (True, False)

2. Define tournamentSelection(Npop, Ntournament): 
    - Get the best Npop individuals from previous generations 
    - Randomly select Ntournament challengers among them
    - Get the best individuals from the challengers
    - Repeat twice to get a father and a mother 
    - return father, mother 
3. Define crossoverMutation(pere.param.value, mere.param.value, param.search_space): 
    - sample alp and mut from a Uniform(0,1) distribution
    - If search_space.log is True then  pere.param.value, mere.param.value = log10(pere.param.value), log10(mere.param.value)
    - Cross-over, and define child.param.value as :
        - Numerical : alp*pere.param.value + (1-alp)*mere.param.value
        - Integer : round( alp*pere.param.value + (1-alp)*mere.param.value)
        - Categorical : random choice (list = [pere.param.value, mere.param.value], prob = [alp, 1-alp]) 
    - Mutation if mut < pmutQuant or mut < pmutCat for quantitative and categorical features respectively and update child.param.value
        - Categorical : Randomly selected an other modality 
        - Integer : Randomly add or substract 1 
        - Numerical : Add sigma * N(0,1) 
    - If search_space.log is True then child.param.value = 10^child.param.value
    - Return child.param.value

4. Define geneticSearch(Npop, Ntournament) :
    - pere, mere = tournamentSelection(Npop, Ntournament)
    - for param in  hyperparameter_list :
        - child.param.value = crossoverMutation(pere.param.value, mere.param.value, param.search_space)
    - return child.param

Step 1 : Initialize genetic algorithm

- Sample Npop individuals (i.e set of hyperparameters) using random search

Step 2 : Run genetic algorithm
- generation = 1
While generation <  Ngeneration :	
    generation = generation + 1
    sample Npop individuals using geneticSearch method

3) Basic package use

@@

3.1) Installation

reservoirnet is an R package api making the python module reservoirPy easily callable from R. It is available on CRAN (see https://cran.r-project.org/package=reservoirnet) and can be installed using:

install.packages("reservoirnet")

3.2) Classification

Vignette:
https://cran.r-project.org/web/packages/reservoirnet/vignettes/Classification_with_RC.html

3.3) Regression

Vignette:
https://cran.r-project.org/web/packages/reservoirnet/vignettes/basic_usage_01.html

4) Study case: Prediction covid-19

@@

<!-- * \proglang{Java} -->
<!-- * \pkg{plyr} -->

@@

<!-- ```{r} -->
<!-- x <- 1:10 -->
<!-- x -->
<!-- ``` -->

@@

<!-- * Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409). -->
<!-- * Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format. -->

>>>>

---
documentclass: jss
author:
  - name: FirstName LastName
    orcid: 0000-0000-0000-0000
    affiliation: University/Company
    # use this syntax to add text on several lines
    address: |
      | First line
      | Second line
    email: \email{name@company.com}
    url: https://posit.co
  - name: Second Author
    orcid: 0000-0000-0000-0000
    affiliation: 'Affiliation \AND'
    # To add another line, use \AND at the end of the previous one as above
  - name: Third Author
    orcid: 0000-0000-0000-0000
    address: |
      | Department of Statistics and Mathematics,
      | Faculty of Biosciences,
      | Universitat Autònoma de Barcelona
    affiliation: |
      | Universitat Autònoma 
      | de Barcelona
    # use a different affiliation in adress field (differently formated here)
    affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "A Capitalized Title: Something about a Package \\pkg{foo}"
  # If you use tex in the formatted title, also supply version without
  plain:     "A Capitalized Title: Something about a Package foo"
  # For running headers, if needed
  short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
  bookdown::word_document2:
    default
knit: (function(inputFile, encoding){
  rmarkdown::render(inputFile, encoding = encoding,
  output_dir = here::here("output"), output_format = "all") })
---

@@

1.  Initialize :

-   Population parameters : $N_{pop} = 200$, $N_e = 100$,
    $N_{generation} = 30$
-   Cross-over parameters : $N_{tournament} = 2$
-   Mutation parameters : $p_{mutQuant} = .5$, $p_{mutCat} = .25$,
    $\sigma = 1$ ($\sigma = 0.1$ for leaking rate as the hyperparameter
    range is shorter)
-   Define the $hyperparameter\_list$, a list containing each
    hyperparameter denoted as $param$.
-   `for` each $param$ in $hyperparameter\_list$ define hyperparameter
    $search\_space$ :
    -   the $lower$ and $upper$ bounds,
    -   the $type$ (numeric, integer, categorical)
    -   if it should be $log$ transformed (True, False)

2.  Define main functions
    -   Define $tournamentSelection(N_{pop}, N_{tournament})$:
        -   Get the best $N_{pop}$ individuals from previous generations
        -   Randomly select $N_{tournament}$ challengers among them
        -   Get the best individuals from the challengers
        -   Repeat twice to get a father and a mother
        -   `return` $father, \: mother$
    -   Define
        $crossoverMutation(pere\$param\$value, mere\$param\$value, param\$search\_space)$:
        -   `if` $search\_space\$log$ $\newline$
            $\text{} \qquad pere\$param\$value = log_{10}(pere\$param\$value)$
            $\newline$
            $\text{} \qquad mere\$param\$value = log_{10}(mere\$param\$value)$

        -   sample $alp$ and $mut$ probabilities from $\mathcal{U}(0,1)$

        -   Cross-over, and define $child\$param\$value$ as :

            -   Numerical :
                $alp*pere\$param\$value + (1-alp)*mere\$param\$value$
            -   Integer :
                $round( alp*pere\$param\$value + (1-alp)*mere\$param\$value)$
            -   Categorical :
                $sample(x = c(pere\$param\$value, mere\$param\$value), \newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad prob = c(alp, 1-alp))$

        -   `if` $mut < p_{mutQuant}$ or $mut < p_{mutCat}$ for
            quantitative and categorical features respectively, mutate :

            -   Categorical : sample another modality
            -   Integer : add or substract 1 at random
            -   Numerical : Add $\sigma \times \mathcal{N}(0,1)$

        -   `if` $search\_space\$log$ :
            $child\$param\$value = 10^{child\$param\$value}$

        -   `return` $child\$param\$value$
    -   Define $geneticSearch(N_{pop}, N_{tournament})$ :
        -   $pere, \: mere = tournamentSelection(N_{pop}, N_{tournament})$
        -   `for` $param$ in $hyperparameter\_list$ : $\newline$
            $\text{} \qquad child\$param\$value = crossoverMutation(\newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad pere\$param\$value, \newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad mere\$param\$value, \newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad param\$search\_space)$
        -   `return` $child$
3.  Run the algorithm
    -   Step 1 : Initialize genetic algorithm
        -   Sample $N_{pop}$ individuals (i.e set of hyperparameters)
            using random search
    -   Step 2 : Run genetic algorithm
        -   $generation = 1$
        -   `while` $generation < Ngeneration$ :
            -   $generation = generation + 1$
            -   sample $N_{pop}$ individuals using $geneticSearch$
                function

# Basic package use

@@

## Installation

reservoirnet is an R package api making the python module reservoirPy
easily callable from R. It is available on CRAN (see
<https://cran.r-project.org/package=reservoirnet>) and can be installed
using:

```{r eval=FALSE}
install.packages("reservoirnet")
```

## Classification

Reservoir Computing (RC) is well suited to both regression and classification tasks. We will introduce a simple example of classification task. This chapter is largely inspired from reservoirnet vignette on
[classification](https://cran.r-project.org/web/packages/reservoirnet/vignettes/Classification_with_RC.html).

The first step is to load useful packages. For this example we will load both reservoirnet and ggplot2 (add ref) a useful package for datavisualisation.

```{r}
library(reservoirnet)
library(ggplot2)
```

### The Japanese vowel dataset

The Japanese vowel dataset is composed of 640 utterances of the Japanese vowel \ae, from 9 different male speakers. The goal of this task is to assign to each utterance the label of its speaker. Dataset is split between a 270 utterances training set and a 340 utterances testing set. Each spoken utterance is a timeseries of 7\~29 timesteps. Each timestep of signal is a 12 dimensional vector representing Linear Prediction Coefficient (LPC), which encode the audio signal into the cepstral domain (a variant of the frequency domain).

References M. Kudo, J. Toyama and M. Shimbo. (1999). "Multidimensional
Curve Classification Using Passing-Through Regions". Pattern Recognition
Letters, Vol. 20, No. 11--13, pages 1103--1111.
<https://archive.ics.uci.edu/ml/machine-learning-databases/JapaneseVowels-mld/>

The data set is splitted into a train and a test set.

```{r}
japanese_vowels <- reservoirnet::generate_data(dataset = "japanese_vowels")$japanese_vowels

X_train <- japanese_vowels$X_train
Y_train <- japanese_vowels$Y_train
X_test <- japanese_vowels$X_test
Y_test <- japanese_vowels$Y_test
```

```{r fig.cap="Vowel dataset"}
sample_per_speaker <- 30
n_speaker <- 9
X_train_per_speaker <- list()

for (i in 0:8) {
 X_speaker <- X_train[((i*sample_per_speaker)+1):((i+1)*sample_per_speaker)]
 X_train_per_speaker[[i+1]]<-(as.numeric(unlist(sapply(X_speaker, t))))
}

d <- data.frame(LPC = unlist(X_train_per_speaker), 
                Speaker = factor(rep(1:9,times = sapply(X_train_per_speaker,length))))

# plot data
ggplot(d,aes(x = Speaker, y = LPC)) + geom_boxplot()+theme_bw()
```

### Echo state network setup

#### Transduction (sequence-to-sequence model)

As ReservoirPy Nodes are built to work on sequences, the simplest setup
to solve this task is sequence-to-sequence encoding, also called
transduction. A model is trained on encoding each vector of input
sequence into a new vector in the output space. Thus, a sequence of
audio yields a sequence of label, one label per timestep.

```{r}
japanese_vowels <- reservoirnet::generate_data(
    dataset = "japanese_vowels",
    repeat_targets=TRUE)$japanese_vowels

X_train <- japanese_vowels$X_train
Y_train <- japanese_vowels$Y_train
X_test <- japanese_vowels$X_test
Y_test <- japanese_vowels$Y_test
```

#### Train a simple Echo State Network to solve this task:

```{r}
source <- createNode("Input")
readout <- createNode("Ridge",ridge=1e-6)
reservoir <- createNode("Reservoir",units = 500,lr=0.1, sr=0.9)

#[source >> reservoir, source] >> readout
model <- list(source %>>% reservoir, source) %>>% readout
```

#### Fit the model

```{r}
model_fit <- reservoirnet::reservoirR_fit(node = model,
                                       X = X_train,
                                       Y = Y_train,
                                       stateful = FALSE,
                                       warmup = 2)

Y_pred <- reservoirnet::predict_seq(node = model_fit$fit,
                                 X = X_test,
                                 stateful = FALSE)
```

Get the scores:

There are 9 speakers, hence the output space is 9-dimensional. The
speaker label is the index of the output neuron with maximum activation.

```{r}
accuracy <- function(pred, truth){
  mean(pred == truth)
}
```

```{r}
Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),1,which.max))
Y_test_class <- sapply(Y_test, FUN = function(x) apply(as.matrix(x),1,which.max))
score <- accuracy(array(unlist(Y_pred_class)), array(unlist(Y_test_class)))

print(paste0("Accuracy: ", round(score * 100,3) ,"%"))
```

    ## [1] "Accuracy: 91.296%"

### Classification (sequence-to-vector model)

We can create a more elaborated model where inference is performed only
once on the whole input sequence. Indeed, we only need to assign one
label to each input sequence. This new setup is known as a
sequence-to-vector model, and this is usually the type of model we refer
to when talking about classification of sequencial patterns.

```{r}
japanese_vowels <- reservoirnet::generate_data(
    dataset = "japanese_vowels")$japanese_vowels

X_train <- japanese_vowels$X_train
Y_train <- japanese_vowels$Y_train
X_test <- japanese_vowels$X_test
Y_test <- japanese_vowels$Y_test
```

```{r}
source <- reservoirnet::createNode("Input")
readout <- reservoirnet::createNode("Ridge",ridge=1e-6)
reservoir <- reservoirnet::createNode("Reservoir", units = 500, lr=0.1, sr=0.9)

#source >> reservoir >> readout
model <- source %>>% reservoir %>>% readout
```

We need to modify the training loop by hand a bit to perform this task:

first, we compute all reservoir states over the input sequence using the
reservoir.run method. then, we gather in a list only the last vector of
the states sequence.

```{r}
states_train = list()
k <- 1
for (x in X_train) {
  states <- reservoirnet::predict_seq(node = reservoir, X = x, reset=TRUE)
  states_train[[k]] <- t(as.matrix(states[nrow(states),]))
  k <- k+1
}
```

We can now train the readout only on the last state vectors. Here,
Y_train is an array storing a single label for each utterance.

```{r}
res <- reservoirnet::reservoirR_fit(readout,X = states_train, Y = Y_train)
summary(res)
```

    ## Parametrs using to fit:
    ##  warmup: 0 ; stateful: FALSE ; reset: FALSE 
    ## results of fitting:
    ## 'Ridge-5': Ridge(ridge=1e-06, input_bias=True, in=500, out=9)

We also modify the inference code using the same method as above:

```{r}
Y_pred <- list()
k <- 1
for (x in X_test) {
  states <- reservoirnet::predict_seq(node = reservoir, X = x, reset=TRUE)
  y <- reservoirnet::predict_seq(node = readout, X = as.array(states[nrow(states),]))
  Y_pred[[k]] <- y
  k <- k+1
}
```

```{r}
Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),1,which.max))
Y_test_class <- sapply(Y_test, FUN = function(x) apply(as.matrix(x),1,which.max))

score <- accuracy(pred = Y_test_class,
                       truth = Y_pred_class)

print(paste0("Accuracy: ", round(score * 100,3) ,"%"))
```

    ## [1] "Accuracy: 88.108%"

## Regression

Vignette:
<https://cran.r-project.org/web/packages/reservoirnet/vignettes/basic_usage_01.html>

# Study case: Prediction covid-19

@@

<!-- * \proglang{Java} -->

<!-- * \pkg{plyr} -->

@@

<!-- ```{r} -->

<!-- x <- 1:10 -->

<!-- x -->

<!-- ``` -->

@@

<!-- * Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409). -->

<!-- * Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format. -->
